Data Imputation as a Query Optimization

### 2016-12-02 paper/algorithm.tex
paper: added join algo sketched, added chained equations algo sketched, added histogram transformation algo

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan is associated with a set of dirty attributes $D$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each base table using the table statistics, which track the number of null values in each column. If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \ C$.  Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$. A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.  Filters do not change the dirty set.

@@

    \Require{$Q$ is a set of query plans and $\Psi$ relates query plans with join predicates.}
    \Function{OptJoin}{$Q, \Phi$}
    \State $Q \gets \emptyset$
    \For{$q \in Q$}

    \EndFor
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for query planning with imputations.}
\end{algorithm}

\begin{verbatim}		
# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
OptJoin(T⁆, join_preds):
	Q = empty
	for t, pred in T:
		S = T \ {t, pred}
		# Get the optimal plans for the left and right arguments to the join.
		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
			if r_l, r_r in join_preds:
				# Add imputation to the left and right arguments.
				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
	return OptRel(Q)
# Extended cost calculation to account for information loss/computation cost in imputations
Cost(q):
  case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
  case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
  case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
  case _ => normal cost calculation

info_loss(op, t, a):
    # beta: some scaling factor
    case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
    case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)

impute_cost(t):
  # alpha: some scaling factor
  alpha * (ntuples(t) + ct(attr(t)))
\end{verbatim}

\subsection{Histogram computation}
Histogram adjustment algorithm:
Invariant: the distribution of non-null values is preserved
Drop(t, ix) => remove nulls from t table statistics for indices ix
Impute(t, ix) => redistribute nulls from t table statistics for indices ix to other buckets based on existing distribution
Filter(t, predicate) => scale all buckets of t table statistics and null counts by predicate selectivity
Join(t1, t2) => estimate cardinality of join result (following selinger, if non-equality, then 30% of product,
If equality and no primary/foreign key relationship then the maximum of the two cardinalities, otherwise
If equality and primary/foreign key relationship then the size ofthe foreign key relationship table). Scale t1's statistics and null counts to total the new cardinality estimate. Do same for t2. Union histograms.

>>>>

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan is associated with a set of dirty attributes $D$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each base table using the table statistics, which track the number of null values in each column. If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \setminus C$.  Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$. A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.  Filters do not change the dirty set.

@@

    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
    \Ensure{Returns a set of optimal query plans for all joins, with distinct dirty sets.}
    \Function{OptJoin}{$Q, \Phi$}
    \For{$size \in 1...|\Psi|$}
    	\State $S \gets subset(\Psi, size)$
	\For{$\phi \in S$}
		\State $S' \gets S \setminus \phi$
			\If{$joins(S', \phi)$}
				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
				\State $(t, \_) \gets \Call{Rels}{\phi}$
				\State $t_{plans} \gets Q(t)$
				\For {$l, r \in t_{plans} \times S'_{plans}$}
					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\phi}}$
					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\phi}}$
					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\phi} p_r | p_l \in P_l, p_r \in P_r \} }$}
				\EndFor
		\EndIf
	\EndFor
    \EndFor
    \State \Return{Q(\Call{Rels}{$\Phi$})}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for query planning with imputations.}
\end{algorithm}

%\begin{verbatim}		
%# For a set of tables (with filter predicates) and join predicates, return the optimal plans.
%OptJoin(T⁆, join_preds):
%	Q = empty
%	for t, pred in T:
%		S = T \ {t, pred}
%		# Get the optimal plans for the left and right arguments to the join.
%		for r_l, r_r in OptJoin(S) x OptFilter(t, pred):
%			if r_l, r_r in join_preds:
%				# Add imputation to the left and right arguments.
%				Q = Q union OptRel(Join(r_l', r_r') for r_l' in AddImpute(r_l, Attrs(join_pred)), r_r' in AddImpute(r_r, Attrs(join_pred)))
%	return OptRel(Q)
%# Extended cost calculation to account for information loss/computation cost in imputations
%Cost(q):
%case drop_min(q', d) =>  info_loss(drop, q', attrs(q') intersect d) + Cost(q')
%case impute_all(q') => info_loss(impute, q', attr(q')) + impute_cost(q') + Cost(q')
%case impute_min(q', d) => info_loss(impute, q', attrs(q') intersect d) + impute_cost(q'[d]) + Cost(q')
%case _ => normal cost calculation
%
%
%info_loss(op, t, a):
%# beta: some scaling factor
%case info_loss(drop, t, a) => beta * sum(for ct_missing(t.a) in a)
%case info_loss(impute, t, a) => beta *sum(for ct_missing(t.a) / ct_complete(t) in a)
%
%impute_cost(t):
%# alpha: some scaling factor
%alpha * (ntuples(t) + ct(attr(t)))
%\end{verbatim}

\subsection{Histogram computation}
In order to correctly estimate the imputation cost and computation time parameters in our cost model,  the system must have updated cardinality estimates
at each point in a query plan. These cardinality estimates can be impacted not just be filtering or joining, as in the traditional relational calculus, but also by
the imputation operators. For example,  dropping versus imputing tuples has a clearly different effect on cardinality. For simplicity of design,
each of the logical nodes in a query plan keeps a reference to a set of histograms. The optimizer creates a copy-on-write of the histograms and modifies them
as necessary to account for new operations in the plan. In all cases, modifications maintain as an invariant that the distribution of data in the new histogram
matches the distribution (but not necessarily the cardinality) of the original statistics. Algorithm~\ref{algo:histogram-transformation} provides the details
of our implementation.

\begin{algorithm}
  \begin{algorithmic}
    \Require{$H$ is a map from attribute names to histograms, $op$ is an operator node in a logical query plan}
    \Ensure{Returns an updated $H$, such that the distribution of data matches the original $H$}

    \Function{UpdateHistogram}{$H, op$}
    	\If{$op \not\in \{\delta, \mu, \sigma, \bowtie \}$}
		\Return{H}
	\EndIf
	\State $H' \gets \Call{Copy}{H}$
    	\If{op = $\delta_C$}
		\For{$c \in C$}
			\State $H'[c][null] \gets 0$
		\EndFor
	\ElsIf{op = $\mu_C$}
		\For{$c \in C$}
			\State \Call{Add}{$H'[c], H'[c][null]$}
		\EndFor
	\ElsIf{op = $\sigma$}
		\State \Call{ScaleBy}{H',  \Call{Selectivity}{$\sigma$}}
	\ElsIf{op = $\bowtie_\phi$}
		\State \Call{ScaleTo}{H', \Call{Cardinality}{$\bowtie_{\phi}$}}
	\EndIf
	\Return{H'}
    \EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
\end{algorithm}

\subsection{Imputation Strategies}
The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT) \todo{CITE}. Note that plugging in a new
imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.

\begin{algorithm}
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCERT}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for in-plan histogram updates}
  \label{algo:histogram-transformation}
\end{algorithm}

\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}
